---
title: "Red State Tutorial"
graphics: yes
output: 
  html_document:
    toc: true
    toc_depth: 1
---

```{r, include = F}
library(knitr)
opts_chunk$set(fig.align = "center", 
               message = F, 
               tidy=TRUE, 
               fig.height = 4, fig.width = 8,
               tidy.opts=list(width.cutoff=50))
```

# Introduction
> I never said all Democrats are saloon-keepers. What I said is that all saloonkeepers
> are Democrats.
>—Horace Greeley, 1860

> Pat doesn’t have a mink coat. But she does have a respectable Republican
> cloth coat.
>—Richard Nixon, 1952

> Like upscale areas everywhere, from Silicon Valley to Chicago’s North
> Shore to suburban Connecticut, Montgomery County supported the
> Democratic ticket in last year’s presidential election, by a margin of 63
> percent to 34 percent.
>—David Brooks, 2001

> There is, for example, this large class of affluent professionals who are solidly
> Democratic. DataQuick Information Systems recently put out a list of 100
> ZIP code areas where the median home price was above $500,000. By my
> count, at least 90 of these places — from the Upper West Side to Santa
> Monica — elect liberal Democrats.
>—David Brooks, 2004

> A lot of Bush’s red zones can be traced to wealthy enclaves or sun-belt
> suburbs where tax cuts are king.
>—Matt Bai, 2001

> But in the Ipsos-Reid surveys, 38% of voters in “strong Bush” counties said
> that they had household incomes below $30,000, while 7% said that their
> families earned at least $100,000. In “strong Gore” counties, by contrast,
> only 29% of voters pegged their household income below $30,000, while
> 14% said that it was above $100,000.
>—James Barnes, 2002

For decades, the Democrats have been viewed as the party of the poor, with the
Republicans representing the rich. Recent presidential elections, however, have
shown a reverse pattern, with Democrats performing well in the richer blue
states in the northeast and coasts, and Republicans dominating in the red states
in the middle of the country and the south.

The purpose of this homework assignment, then, is to use Stan and hierarchical
models to understand this "paradox". For our purposes, we will use data from the
2004 election.

# Preliminaries

If you haven't already, please install RStan. Instructions can be found here:
<https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started>.

Then, please download the folder with the data to your desktop. In your R
session, you'll need to set the working directory to this folder.

For Windows, the command to do so will be:
```{r eval = F}
setwd("C:/Users/Public/Desktop/Red State Tutorial")
```

For Macs, the command will be:
```{r eval = F}
setwd("~/Desktop/Red State Tutorial")
```

In addition, you should also run the following commands in your R Session, which
will load RStan, set it up to take full advantage of your computer's
processing power, and load some helper functions that we'll use later.

```{r results = 'hide'}
pkgs <- c("car", "dplyr", "ggplot2", "reshape2")
inst <- which(!pkgs %in% installed.packages()[,"Package"])
if (length(inst)) install.packages(pkgs[inst])

library(rstan)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
source("red_state_tutorial_helper_functions.R")
```


#Visualizing the "Paradox"

Let's first try to analyze the data and see if this "paradox" exists. There are
two parts. The first is that the Democrats are the "party of the poor". What do
we mean when we say that? Well, we expect that people with lower income are more
likely to vote Democrat and people with higher income are more likely to vote
Republican. Is this the case?

We'll use data from the National Annenberg 2004 Election Survey to
explore whether income is positively correlated with voting Republican. In
this data set, income is not a continuous variable, but divided into five
categories: \$0 - \$20,000, \$20,000 - \$40,000, \$40,000 - \$75,000, \$75,000 -
\$150,000, and \$150,000+. Looking at the data, we see:

```{r Annenberg data plot, echo = F, fig.width = 9, fig.height = 4}
library(car)
library(dplyr)
library(ggplot2)

#Set default theme for ggplot2
theme_set(theme_classic())

#Build individual data
dat.vot <- read.table("Data/votechoice2000-04-08.dat", header=T, sep="\t")
dat.vot$weight[is.na(dat.vot$weight)] <- 1
ok <- apply(is.na(dat.vot[1:8]), MARGIN = 1, FUN = sum) == 0 & 
  (dat.vot$cit == 1 | is.na(dat.vot$cit)) & 
  (dat.vot$regist == "registered" | is.na(dat.vot$regist))
dat.vot <- subset(dat.vot, ok)
recodes <- "'2000-ann'=2000; '2004-ann'=2004; '2008-pew'=2008"
dat.vot$year <- recode(dat.vot$file, recodes, as.factor.result = FALSE)
dat.vot <- subset(dat.vot, year != 2000)

plot_data <- summarize(group_by(dat.vot[dat.vot$year == 2004,], inc),
                       num_votes = n(), rep_votes = mean(rvote), 
                       dem_votes = 1 - rep_votes)
graph_data <- ggplot(plot_data, aes(x = inc, y = rep_votes)) + 
  geom_line(aes(color = "Rep"), size = 1) + 
  geom_line(aes(y = dem_votes, color = "Dem"), size = 1) + 
  geom_point(aes(color = "Rep"), size = 4) + 
  geom_point(aes(y = dem_votes, color = "Dem"), size = 4) + 
  geom_bar(aes(y = num_votes/sum(num_votes), fill = "All"), stat = "identity") +
  scale_color_manual(name = "Party", values = c(Rep = "red", Dem = "blue"),
                     guide = guide_legend(order = 1)) +
  scale_fill_manual(name = "", values = "gray35", 
                    labels = "Proportion of \ntotal votes from \nincome category",
                    guide = guide_legend(order = 2)) +
  labs(x = "Income Level", y = "Proportion of votes") + 
  ggtitle("Voting by Income Level")

graph_data
```

The red and blue points in the plot show the proportion of the respondents in
each income category who voted Republican and Democrat, respectively. The bars
below show the proportion of the respondents in each income category (i.e. the 
distribution of income level in the sample). It looks like the proportion of 
Republican votes increases with each income level. 

For simplicity we're ignoring third party voters, which means that we can 
make a binary variable that is 0 if the respondent voted Democrat and 1 if 
they voted Republican. We can then do a logistic regression with 
with income level as the predictor of voting Republican. 

Because income is a categorical variable with five levels, we estimate a 
regression coefficient for four of the levels and interpret the coefficients 
relative to the omitted reference level. The intercept of the model can then
also be interpreted as the coefficient on the lowest level (alternatively we can
think of it as omitting the intercept and estimating a coefficient for all five
of the levels). We're trying to see if the likelihood of voting Republican
increases when income increases, so it makes sense to use the first (i.e.
lowest) income level as the reference level.

Now that we know our model, we can use Stan to estimate it. 
For this example, the Stan code and data for the model are
provided, so all you have to do is run the model:

```{r Individual Income Stan, results = "hide"}
load("Data/stan_inc_regression_data.Rdata")

#Use Stan to run the regression model
model1_file <- "Stan/regression_by_one_var.stan"
model1_name <- "Republican vote by individual income"
model1 <- stan(file = model1_file, data = stan_2004_ind_inc_data, 
               model_name = model1_name)
```

The `stan` function call returns a `stanfit` object. To read more about it, you 
can visit <http://mc-stan.org/interfaces/rstan.html> and click on the RStan 
Manual and Vignette link.

In the Stan code, the regression coefficients for income level are named 
`b_inc2`, `b_inc3`, `b_inc4`, and `b_inc5`, one for each level between two and 
five. We can use the `summary` function to pull out the posterior means and 95% 
credible intervals for the coefficients. Credible intervals are similar to
confidence intervals, but they have a more intuitive interpretation in terms of
probabilities. Conditional on our model/data/assumptions, we can say that the
probability that the paramater is in the $100p\%$ credible interval is $p$. To
get the 95% intervals we pull out the 2.5% and 97.5% quantiles.

(The mean is just one possible point estimate to look at. Another common choice
is the median. We could also look at other credible intervals. Point estimates
and credible intervals are just summaries that help us make inferences and learn
from the model, so we can look at whichever ones we think will help us do that.)

```{r Individual Income Plot Results, fig.width=4}
# Get the means and 95% intervals 
coef_nms <- c("Intercept", "b_inc2", "b_inc3", "b_inc4", "b_inc5")
stat_nms <- c("mean", "2.5%", "97.5%")
plot_data <- make_plot_data(model1, coef_nms, stat_nms)
# Plot the estimates
plot_estimates(plot_data) + 
      ggtitle("Individual Income Level \ncoefficients & 95% intervals")
```

You'll notice in the plot that the minimum value is a bit less than $-0.5$ and 
the largest is a bit less than 1. Clearly, the coefficients do not represent 
probabilities because probabilities must be between 0 and 1. Because of the 
transformation used for logistic regression the coefficients are on the log-odds
(or logit) scale. The inverse logit function (plotted below) is used to
transform real numbers (the x-axis in the plot) to probabilities between 0 and 1
(the y-axis):

```{r inv logit function, echo = F}
(graph_invlogit <- 
  ggplot(plot_data, aes(x = mean, y = plogis(mean), 
                        fill = factor(mean), color = factor(mean))) + 
  geom_segment(aes(x = -Inf, xend = mean, y = plogis(mean), yend = plogis(mean)), 
               linetype = 2, size = 0.25, show_guide = FALSE) + 
  geom_segment(aes(x = mean, xend = mean, y = -Inf, yend = plogis(mean)), 
               linetype = 2, size = 0.25, show_guide = FALSE) + 
  stat_function(fun = plogis, color = "gray35", size = 1.5) + 
  geom_point(shape = 21, size = 5, color = "gray35") + 
  scale_x_continuous(limits = c(-10, 10)) + 
  scale_fill_discrete("Coefficient", labels = plot_data$name) +
  labs(x = "x", y = "inverse-logit(x)") + 
  ggtitle("Inverse Logit Function"))
```

As we can see in the plot, larger coefficients (larger x values) mean higher
probabilities of voting Republican though how much of a difference depends on
which two points we compare. So it does look like an increase in an individual's
income is associated with an increase in the probability of voting Republican,
which supports that idea that the Democrats are the party of the poor.

However, the other part of the paradox is that the Democrats are the party of 
the rich states. In other words, are states with higher average income more 
likely to vote Democrat on average? We can again use a logistic regression model,
however instead of using the 2004 Annenberg Survey data we'll be using the
Census CPS data. Thus, while our state average income level variable will still
have five categories, we'll divide it so that there are the same number of
states in each category. The categories are: \$0 - \$39,000, \$39,000 -
\$43,003, \$43,003 - \$44,476, \$44,476 - \$50,614 and \$50,614+.

Our model is the same as before so we can reuse the same Stan code and just 
give Stan the state-level data instead of the individual-level data.

```{r State Income Stan, results = "hide"}
# Load the state-level data
load("Data/stan_state_regression_data.Rdata")

# Fit the model using Stan
model2_name <- "Republican vote by state average income"
model2 <- stan(fit = model1, data = stan_2004_state_inc_data, 
               model_name = model2_name)
```

We'll again summarize the data in the same way as before. Only this time, we'll be doing it using a function instead of looking at the Stanfit object itself.

And now, we'll plot it, but we'll again use a function that's wrapping the code called previously.

```{R State Income Plot, fig.width=4, results="hold"}
plot_data2 <- make_plot_data(model2, coef_nms, stat_nms)
plot_estimates(plot_data2) + 
  ggtitle("State Income Level \ncoefficients & 95% intervals")

# Note: we can't really see the 95% intervals in this plot because they are much
# narrower for the coefficients on state average income than individual income.
```

In our plot, we see that as income level increases, the coefficients decrease.
Based on our understanding of the inverse logit function, this means the
probability of voting Republican decreases as well. 

All in all, we can see that as an individual's income increases, the likelihood
of voting Republican does so as well, but as a state's average income increases,
the likelihood of voting Republican goes down. Let's now try to understand the
paradox.


# Understanding this "paradox"

For our first attempt to understand this paradox, what if we treated the state's
average income level and individual's income level as predictor variables and
added their effects together? Our outcome is still the same as before so we'll
use a binomial logistic regression as before. On the other hand, we now have two
categorical variables. Still, for each variable, we'll compare the outcome of
one level to those of all other levels. So, let's run the Stan code.

```{r Ind and State Income Stan, results = "hide"}
# Load the data
load("Data/stan_state_inc_and_ind_inc_regression_data.Rdata")

# Fit the model using Stan
model3_file <- "Stan/regression_by_state_inc_and_ind_inc.stan"
model3_name <- "Republican vote by state and ind income"
model3 <- stan(file = model3_file, data = stan_2004_state_and_ind_inc_data, 
               model_name = model3_name)
```

For the model individual income level regression coefficients are  `b_ind_inc2`,
`b_ind_inc3`, `b_ind_inc4`, and `b_ind_inc5`, and our state average income level
regression coefficients are `b_state_inc2`, `b_state_inc3`, `b_state_inc4`, and
`b_state_inc5`. We don't have regression coefficients for the first or lowest
levels because again, we've set our model to use the lowest level as the
reference categories for comparisons. However, the intercept now represents the
effect of the lowest individual income level and lowest state average income
level on voting Republican.

Let's plot these estimates:
```{r Ind and State Income Plot}
coef_nms <- c("Intercept", "b_ind_inc2", "b_ind_inc3", "b_ind_inc4", "b_ind_inc5",
              "b_state_inc2", "b_state_inc3", "b_state_inc4", "b_state_inc5")
plot_data3 <- make_plot_data(model3, coef_nms, stat_nms)
plot_estimates(plot_data3) + 
  ggtitle("Estimated coefficients for state and individual income level")
```

It looks like the regression coefficients for individual income level increase
as income level increases whereas the regression coefficients for state income
level decrease as income level increases. Because our model is linear on the
logit scale, one way to interpret these regression coefficients is that:

1. For a given individual income level, the likelihood of voting Republican
decreases as the state average income increases.

2. For a given state average income level, the likelihood of voting Republican
increases as the state average income increases.

So, we haven't really gotten any more insight into the "paradox". 

Let's take a step back and think about this problem some more. What do we know?

1. We know that some states are more likely to vote Republican.

2. We know that individuals with higher income are more likely to vote Republican.

3. We know that states with higher average income are less likely to vote Republican.

To address point 1 we can allow the intercept in our model to vary by state. 
After all, to say that some states are more likely to vote Republican, we're
also essentially saying that we want the intercept in our regression to be able
to be different for each state. 

And if we can vary the intercept by state, why not also vary the slope
associated with an individual's income by state? (It doesn't really make sense
to vary the slope associated with a state's average income by state.) In other
words, are all individuals with higher income more likely to vote Republican or
does this vary by state?

Let's conduct this analysis. Our outcome will still be the same so we'll still
use a binomial logistic regression model.

```{r Ind Income by state Stan, results = "hide"}
load("Data/stan_state_and_inc_regression_data.Rdata")
model4_file <- "Stan/mixed_effects_regression_for_state_and_inc.stan"
model4_name <- "Republican vote by income and state"
model4 <- stan(file = model4_file, data = stan_2004_state_and_income_data,
               model_name = model4_name)
```

From this model we obtain estimates for each of the states. Let's focus on
Mississippi, Ohio, and Conneticut because they represent a poor, middle-income,
and rich state, respectively.

```{r Ind Income by state Plot}
# Pull out the means and intervals
state_nums <- c(MS = 25, OH = 34, CT = 6)
coef_ct <- generate_state_labels(state_nums["CT"])
coef_oh <- generate_state_labels(state_nums["OH"])
coef_ms <- generate_state_labels(state_nums["MS"])
plot_data4 <- rbind(make_plot_data(model4, coef_ms, stat_nms), 
                    make_plot_data(model4, coef_oh, stat_nms), 
                    make_plot_data(model4, coef_ct, stat_nms))
# Add variables indicating the state and income level
plot_data4$state <- gl(3, 5, labels = names(state_nums))
plot_data4$inc <- 1:5

# Plot the estimates for the three states side by side
ggplot(plot_data4, aes(x = inc, y = mean, ymin = lower95, 
                       ymax = upper95, color = state, shape = state)) +
  geom_pointrange(size = 1) + 
  geom_hline(yintercept = 0, linetype = 2, color = "maroon") +
  labs(y = "Estimate", x = "Income Level") +
  ggtitle("Coefficient Estimates: means and 95% intervals\n") + 
  facet_wrap(~state, scales = "fixed")
``` 

Although individuals with higher incomes are more likely to vote Republican, 
this trend is less extreme for Connecticut than in Ohio and less extreme for
Ohio than Mississippi. 

Let's take a step back and see what we've learned. We sought to understand the 
paradox that even though the Democrats claim to be the party of the poor, states
with higher average income tended to swing Democrat. When we analyzed the 2004 
National Election Annenberg Survey data using a logistic regression on 
individual income level and the number of Republican voters, we did indeed see 
that the likelihood of voting Republican increases as income level increases. On
the other hand, when we analyzed the Census CPS data using a logistic regression
on state average income level and the Republican vote share, we saw that the
likelihood of voting Republican decreases as average income level increases. So,
we confirmed that the paradox seemed to exist.

When we sought to decipher the paradox, we saw that using individual income 
level and state average income level as predictors model didn't help. Instead, 
it was through a hierarchical regression model in which we allowed the intercept
and individual income to vary by state. In our plots, we saw that while
individuals with higher income are more likely to vote Republican, the effect is
smaller in states with higher average income. This explained our paradox.

So, if you enjoy doing this analysis and would like to see this analysis done on
other elections, you can read Andrew Gelman, Boris Shor, Joseph Bafumi, and
David Park's 
*Rich State, Poor State, Red State, Blue State: What’s the Matter with Connecticut?* 
(from which this tutorial is derived). Or, you could do it yourself and compare your 
results to their paper.
